Using algorithmic skeletons in EcmaScript to parallelize computation in browsers through Web Workers

### 2015-05-16 docs/paper/paper.tex

\section{Background}\label{sec:background}
All modern versions of major browsers (Internet Explorer, Mozilla Firefox, Google Chrome, Safari and Opera) allow the execution of JavaScript code in a single threaded event loop runtime \cite{event-loop}. User actions add elements to the event loop queue and these are processesed sequentially. All process input/output is asynchronous to prevent the UI thread from blocking, thus keeping it responsive.
While such a runtime is a good fit for most web applications, JavaScript and the browser are now also being used for some CPU intensive tasks, such as games, which commonly require physics calculations, image manipulation (in 2D) and graphics generation (in 3D). Despite the large improvement in JavaScript performance obtained through optimizing compilers such as Google Chrome's V8 \cite{v8} and Mozilla Firefox's SpiderMonkey \cite{spider-monkey}, it is important to be able to take advantage of multiple cores in modern devices to achieve even greater performance.\\
In 2010, Web Workers were made part of the web standard. Web Workers allow the creation of ``thread like'' constructs in a browser environment, but they don't allow on shared memory and instead communicate via message passing. The message passing overhead is very big for common objects and Web Workers ``have a high start-up performance cost'' \cite{w3c-ww}, so they were commonly considered for long running tasks and operating on generally the same set of data during a single execution, making them unfit for a thread pool  model \cite{thread-pool}.\\
However, a spec is being developed for version 7 of the EcmaScript standard that changes this, by introducing the notion of shared memory through \verb+SharedArrayBuffer+s \cite{sab}. In essence, the proposal allows the same memory to be shared across multiple Web Workers, and also aims to provide the necessary atomic/lock constructs to deal with shared memory. In the particular case of embarrassingly parallel computations (such as map, filter and reduce operations on an array), it is simple to take advantage of the shared memory speed benefits without incurring in any overhead due to locks.

@@

To verify our hypothesis, we put together two benchmarks\footnote{All benchmarks in this document use the environment described in Tab.~\ref{tab:env}.} that aim to verify the difference between invoking \verb+postMessage+ with and without structured cloning for a \verb+SharedTypedArray+. Both transfer a \verb+SharedTypedArray+ back and forth between the UI thread and a worker; the only difference between the two is that one\footnote{http://jsperf.com/transferrable-vs-cloning/2} has 100,000 (a hundred thousand) elements in the \verb+SharedTypedArray+ and the other one\footnote{http://jsperf.com/longer-transferrable-vs-cloning/2} 1,000,000 (a million). As it can be seen from the  results in Tab.~\ref{tab:transf-vs-clone}, increasing the amount of elements by 10x does not change the amount of operations that can be performed when using \verb+Transferable+ objects, it scales. On the other hand, the amount of operations that can be performed with structured cloning greatly decreases.

For that reason our library only works with \verb+SharedTypedArray+s and \verb+TypedArrays+.

@@

We created benchmarks\footnote{http://jsperf.com/pjs-serialization-long/3 and http://jsperf.com/pjs-serialization/3} using two functions whose string representation has different lengths to see if the function's length affected the serialization and transfer time (results in Tab.~\ref{tab:send-function}). We measured three approaches to transfer the string:
\begin{enumerate}
  \item Encoding the \verb+String+ into an \verb+TypedArrayBuffer+, sending it as a \verb+Tranferable+ and decoding it on the Web Worker.
  \item Sending the \verb+String+ directly.
  \item Using a \verb+Blob+ to store the stream and retrieving that in the Worker.
\end{enumerate}

@@

We also created a benchmark\footnote{http://jsperf.com/pjs-encoding/2} that tranfers a \verb+SharedTypedArray+ and a \verb+String+ that are properties of the same \verb+message+ object back and forth between the UI thread an a Web Worker to understand if transferring additional objects affected the transfer time when comparing it with just tranferring a \verb+String+ (results in Tab.~\ref{tab:send-function-and-array}.
\begin{table}
  \centering
  \begin{tabular}{|c|c|}
    \hline
    Copying code [ops/sec] & Tranferring code [ops/sec] \\
    \hline
    \textbf{4509} & 4270 \\
    \hline
  \end{tabular}
  \caption{Sending code and a TypedArray to a Web Worker.}
  \label{tab:send-function-and-array}
\end{table}

@@

We intend to perform some tests to evaluate the possibility of serializing functions via \verb+SharedArrayBuffer+s. We have not done this up to now due to time constraints, but we believe that performance improvements could be achieved by serializing function string once, creating an \verb+SharedTypedArray+ for the bytes and then sharing the backing buffer with all workers.\\

>>>>

\section{Background}\label{sec:background}
All modern versions of major browsers (Internet Explorer, Mozilla Firefox, Google Chrome, Safari and Opera) allow the execution of JavaScript code in a single threaded event loop runtime \cite{event-loop}. User actions add elements to the event loop queue and these are processesed sequentially. All process input/output is asynchronous to prevent the UI thread from blocking, thus keeping it responsive.
While such a runtime is a good fit for most web applications, JavaScript and the browser are now also being used for some CPU intensive tasks, such as games, which commonly require physics calculations, image manipulation (in 2D) and graphics generation (in 3D). Despite the large improvement in JavaScript performance obtained through optimizing compilers such as Google Chrome's V8 \cite{v8} and Mozilla Firefox's SpiderMonkey \cite{spider-monkey}, it is important to be able to take advantage of multiple cores in modern devices to achieve even greater performance.\\
In 2010, Web Workers were made part of the web standard. Web Workers allow the creation of ``thread like'' constructs in a browser environment, but they don't allow on shared memory and instead communicate via message passing. The message passing overhead is very big for common objects and Web Workers ``have a high start-up performance cost'' \cite{w3c-ww}, so they were commonly considered for long running tasks and operating on generally the same set of data during a single execution, making them unfit for a thread pool  model \cite{thread-pool}.\\
However, a spec is being developed for version 7 of the EcmaScript standard that changes this, by introducing shared memory through \verb+SharedArrayBuffer+s \cite{sab}. In essence, the proposal allows the same memory to be shared across multiple Web Workers, and also aims to provide the necessary atomic/lock constructs to deal with shared memory. In the particular case of embarrassingly parallel computations (such as map, filter and reduce operations on an array), it is simple to take advantage of the shared memory speed benefits without incurring in any overhead due to locks.

@@

To verify our hypothesis, we put together two benchmarks\footnote{All benchmarks in this document use the environment described in Tab.~\ref{tab:env}.} that aim to verify the difference between invoking \verb+postMessage+ with and without structured cloning for a \verb+SharedTypedArray+. Both transfer a \verb+SharedTypedArray+ back and forth between the UI thread and a worker; the only difference between the two is that one\footnote{http://jsperf.com/transferrable-vs-cloning/2} has 100,000 (a hundred thousand) elements in the \verb+SharedTypedArray+ and the other one\footnote{http://jsperf.com/longer-transferrable-vs-cloning/2} 1,000,000 (a million). As it can be seen from the  results in Tab.~\ref{tab:transf-vs-clone}, increasing the amount of elements by 10x does not change the amount of operations that can be performed when using \verb+Transferable+ objects, it scales. On the other hand, the amount of operations that can be performed with structured cloning greatly decreases. For that reason our library only works with \verb+SharedTypedArray+s and \verb+TypedArrays+.

@@

We created benchmarks\footnote{http://jsperf.com/pjs-serialization-long/3 and http://jsperf.com/pjs-serialization/3} using two functions whose string representation has different lengths to see if the function's length affected the serialization and transfer time (results in Tab.~\ref{tab:send-function}). We measured three approaches to transfer the string:
\begin{enumerate}
  \item Encoding the \verb+String+ into an \verb+TypedArray+, sending it as a \verb+Tranferable+ and decoding it on the Web Worker.
  \item Sending the \verb+String+ directly.
  \item Using a \verb+Blob+ to store the stream and retrieving that in the Worker.
\end{enumerate}

@@

We created a separate benchmark\footnote{http://jsperf.com/shared-function-transfer} to compare the best option (copying) against encoding the function into a \verb+SharedTypedArray+ and transferring it to four workers. We expected the latter to be slower since it involves encoding the code once and decoding it in each worker. The results are shown in Tab.~\ref{tab:copy-vs-shared}.

\begin{table}
  \centering
  \begin{tabular}{|c|c|}
    \hline
    Copying [ops/sec] & SharedArrayBuffer [ops/sec] \\
    \hline
    \textbf{1315} & 1226 \\
    \hline
  \end{tabular}
  \caption{Sending code to four workers by copying compared to encoding it and using a SharedArrayBuffer.}
  \label{tab:copy-vs-shared}
\end{table}

We also created a benchmark\footnote{http://jsperf.com/pjs-encoding/2} that tranfers a \verb+SharedTypedArray+ and a \verb+String+ that are properties of the same \verb+message+ object back and forth between the UI thread an a Web Worker to understand if transferring additional objects affected the transfer time when comparing it with just tranferring a \verb+String+ (results in Tab.~\ref{tab:send-function-and-array}).
\begin{table}
  \centering
  \begin{tabular}{|c|c|}
    \hline
    Copying code [ops/sec] & Tranferring code [ops/sec] \\
    \hline
    \textbf{4509} & 4270 \\
    \hline
  \end{tabular}
  \caption{Sending code and a TypedArray to a Web Worker.}
  \label{tab:send-function-and-array}
\end{table}

@@
